\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-10 paper/ms.tex
adding carma benchmark figure

We run \exampleiiinburn~steps of burn-in, followed by \exampleiiinsteps~steps
of MCMC using \emcee.
We estimate the integrated autocorrelation time of the chain for $\ln
P_\mathrm{rot}$ and estimate that we have \exampleiiineff\ independent
samples.
These samples give a posterior constraint on the period of $P_\mathrm{rot} =
\rotationperiod$ days and the marginalized posterior distribution for $P$ is shown
in the right panel of \Figure{rotation}.
This result is in good agreement with the literature value with smaller
uncertainties.
A detailed comparison of GP rotation period measurements and the traditional
methods is beyond the scope of this paper, but forthcoming work (R.~Angus,
\etal, submitted) demonstrates that GP inferences are, at a population level,
more reliable than other methods.
}

@@

\response{
In this section, we demonstrated five potential use cases for \celerite.
These examples span a range of data sizes and model complexities, and the
last three are all based on active areas of research with immediate real-world
applicability.
The sizes of these datasets are modest on the scale of some existing and
forthcoming survey datasets because we designed the experiments to be easily
reproducible but, even in these cases, the inferences made here would be
intractable without substantial computational investment.

Table~\ref{tab:example-stats} lists the specifications of each example.

In Example~3, for example, we inferred the rotation period of a star observed
by \kepler\ with a cost of about 4~CPU-minutes.
By contrast, the same inference using a general but optimized Cholesky
factorization routine would require nearly 400~CPU-hours.
The upcoming \tess\ Mission will observe \dfmtodo{XXX} stars.
}

@@

\subsection{Kalman filtering \& CARMA models} \sectlabel{kalman}

\response{
The continuous autoregressive moving average (CARMA) models introduced into
the astrophysics literature by \citet{Kelly:2014} share many features with
\celerite.
CARMA models are derived by solving a stochastic differential equation and,
like \celerite, the likelihood function for a CARMA model conditioned on a
one-dimensional dataset can be solved in $\mathcal{O}(N)$.
The kernel function for a CARMA$(J,\,K)$ model is \citep{Kelly:2014}
\begin{eqnarray}\eqlabel{carma-kernel}
k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{carma-coeff}
A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
\end{eqnarray}
and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
the model; $\sigma$ is real, while the others are complex.
Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
every CARMA model corresponds to an equivalent \celeriteterm\ model and the
parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
The inverse statement is not as simple.
In practice, this means that \celerite\ could be trivially used to compute any
CARMA model.
Using the CARMA solver to compute a \celeriteterm\ model, however, requires
solving \eq{carma-coeff} numerically for a given set of $\{A_j\}_{j=1}^J$.
The computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$ using
the Kalman filtering based method for solving CARMA models \citep{Kelly:2014};
however, in practice this method is somewhat slower.
An additional benefit of the \celeriteterm\ model is that it is parameterized in terms
of the amplitudes directly.
This means that the physical interpretation of these parameters is, in some
cases, simpler.
On the other hand, CARMA models are parameterized by modeling the power
spectrum of the stochastic driving force and this might be the quantity of
interest in other situations.
Another benefit of the \celeriteterm\ framework is that, unlike
filtering-based CARMA solvers, it directly solves matrix equations of the form
${K_\bvec{\alpha}}^{-1}\,b$ in $\mathcal{O}(N)$.
This is a crucial feature when GP modeling is combined with linear models
\citep[for example][]{Luger:2017}.
}

>>>>

We run \exampleiiinburn~steps of burn-in, followed by \exampleiiinsteps~steps
of MCMC using \emcee.
We estimate the integrated autocorrelation time of the chain for $\ln
P_\mathrm{rot}$ and estimate that we have \exampleiiineff\ independent
samples.
These samples give a posterior constraint on the period of $P_\mathrm{rot} =
\rotationperiod$ days and the marginalized posterior distribution for $P$ is
shown in the right panel of \Figure{rotation}.
This result is in good agreement with the literature value with smaller
uncertainties.
A detailed comparison of GP rotation period measurements and the traditional
methods is beyond the scope of this paper, but forthcoming work (R.~Angus,
\etal, submitted) demonstrates that GP inferences are, at a population level,
more reliable than other methods.

The total computational cost for this inference using \celerite\ is about
4~CPU-minutes.
By contrast, the same inference using a general but optimized Cholesky
factorization routine would require nearly 400~CPU-hours.
This speed-up enables probabilistic measurement of rotation periods using
existing data from \project{K2} and forthcoming surveys like \tess\ and \lsst\
where this inference will need to be run for at least hundreds of thousands of
stars.
}

@@

\response{In the previous subsections, we demonstrate five potential use
cases for \celerite.
These examples span a range of data sizes and model complexities, and the last
three are based on active areas of research with immediate real-world
applicability.
The sizes of these datasets are modest on the scale of some existing and
forthcoming survey datasets because we designed the experiments to be easily
reproducible but, even in these cases, the inferences made here would be
intractable without substantial computational investment.

Table~\ref{tab:example-stats} lists the specifications of each example.
In this table, we list the computational cost required for a single likelihood
evaluation using a general Cholesky factorization\footnote{We use the Intel
Math Kernel Library \url{https://software.intel.com/en-us/intel-mkl}}
and the cost of computing the same likelihood using \celerite.
Combining this with the total number of model evaluations required to run the
MCMC algorithm to convergence (this number is also listed in
Table~\ref{tab:example-stats}), we can estimate the total CPU time required in
each case.
While the exact computational requirements for any problem also depend on the
choice of inference algorithm and the specific precision requirements, this
table provides a rough estimate of what to expect for projects at these
scales.
}

@@

The continuous autoregressive moving average (CARMA) models introduced into
the astrophysics literature by \citet{Kelly:2014} share many features with
\celerite.
CARMA models are restricted to one-dimensional problems and the likelihood
function for a CARMA model can be solved in $\mathcal{O}(N\,J^2)$ using a
Kalman filter \citep{Kelly:2014}.
The kernel function for a CARMA$(J,\,K)$ model is
\begin{eqnarray}\eqlabel{carma-kernel}
k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{carma-coeff}
A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
\end{eqnarray}
and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
the model; $\sigma$ is real, while the others are complex.
Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
every CARMA model corresponds to an equivalent \celeriteterm\ model and the
parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
The inverse statement is not as simple.
In practice, this means that \celerite\ could be trivially used to compute any
CARMA model.
Using the CARMA solver to compute a \celeriteterm\ model, however, requires
solving \eq{carma-coeff} numerically for a given set of $\{A_j\}_{j=1}^J$.
The computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$ using
the Kalman filtering based method for solving CARMA models \citep{Kelly:2014};
however, in practice this method is somewhat slower.
An additional benefit of the \celeriteterm\ model is that it is parameterized in terms
of the amplitudes directly.
This means that the physical interpretation of these parameters is, in some
cases, simpler.
On the other hand, CARMA models are parameterized by modeling the power
spectrum of the stochastic driving force and this might be the quantity of
interest in other situations.
Another benefit of the \celeriteterm\ framework is that, unlike
filtering-based CARMA solvers, it directly solves matrix equations of the form
${K_\bvec{\alpha}}^{-1}\,b$ in $\mathcal{O}(N)$.
This is a crucial feature when GP modeling is combined with linear models
\citep[for example][]{Luger:2017}.

\begin{figure}[tp]
\begin{center}
\includegraphics[width=\textwidth]{figures/benchmark_darwin_carma.pdf}
\caption{The same as \Figure{benchmark}, but using an optimized
    \project{C++} implementation of the Kalman filter method developed by
    \citet{Kelly:2014}.
    Comparing this figure to \Figure{benchmark}, we see that \celerite\ is
    about an order of magnitude faster in all cases.
    \emph{(left)} The cost of computing \eq{gp-likelihood} with a covariance
    matrix given by \eq{celerite-kernel} as a function of the number of data
    points $N$.
    The different lines show the cost for different numbers of terms $J$
    increasing from bottom to top.
    To guide the eye, the straight black line without points
    shows linear scaling in $N$.
    \emph{(right)} The same information plotted as a function of $J$ for
    different values of $N$.
    Each line shows the scaling for a specific value of $N$ increasing from
    bottom to top.
    The black line shows quadratic scaling in $J$.
    \figurelabel{benchmark-carma}}
\end{center}
\end{figure}
